Natural selection in the Health and Retirement Study

### 2023-11-30 hrs-selection.qmd
Further data cleaning

# TE this is average number of kids that your kids have
rand$div_h10gkid <- rand$h10gkid/rand$raevbrn
# TE 283 people report having grandkids, but having had no children. Weird!
gkids_unusable_white <- sum(na.omit(ifelse(subset(rand, ethnicity == "white")$h10gkid > 0 & subset(rand, ethnicity == "white")$raevbrn == 0, 1, 0)))

gkids_unusable_black <- sum(na.omit(ifelse(subset(rand, ethnicity == "black")$h10gkid > 0 & subset(rand, ethnicity == "black")$raevbrn==0,1,0)))

# If grandkids but no kids div_gkid is infinite. Replacing with NA
rand$div_h10gkid[is.infinite(rand$div_h10gkid)] <- NA

# TE added some variables for grandkid analysis
# TE I've also added sampling weights in calculation of mean values
# note the child weight is just weight* number of kids. so mean_h10gkid 
# and mean_div_h10gkid are actually the same
rand <- rand |>
  mutate(.by = rabyear, # calculate this pooling ethnicity
    mean_weight = mean(weight, na.rm = TRUE),
    mean_child_weight = mean(child_weight, na.rm = TRUE),
    mean_raevbrn = mean(raevbrn*weight/mean_weight, na.rm = TRUE),
    rlrs = raevbrn/mean_raevbrn,
    mean_h10gkid = mean(h10gkid*weight/mean_weight, na.rm = TRUE),
    rlrs_h10gkid = h10gkid/mean_h10gkid,
    mean_div_h10gkid = mean(div_h10gkid*child_weight/mean_child_weight, na.rm = TRUE), 
    rlrs_div_h10gkid = div_h10gkid/mean_div_h10gkid
  ) |>
  mutate(.by = ethnicity, # calculate these stats within-ethnicity
    income_resid = calc_income(rand, ethnicity[1]),
    income_resid = income_resid - min(income_resid, na.rm = TRUE) + 1,
    income_med = chop_equally(income_resid, 2, labels = c("Low", "High")),
    # as.numeric avoids haven.labelled issues:
    raedyrs = as.numeric(raedyrs),
    r10hearte = as.numeric(r10hearte),
    r10diabe = as.numeric(r10diabe),
    edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")),
    born = chop(rabyear, 1942, labels = c("pre-1942", "post-1942")),
    # wave 10 is 2010
    # table(rands$white$r10mstat)
     # 1                 1.married
     # 2   2.married,spouse absent
     # 3               3.partnered
     # 4               4.separated
     # 5                5.divorced
     # 6      6.separated/divorced
     # 7                 7.widowed
     # 8           8.never married
    married = factor(r10mstat == 1, 
                     levels = c(FALSE, TRUE),
                     labels = c("Other", "Married")),
    r10sibs = r10livsib + 1, # n. sibs including oneself!
    agefbn = r10agey_e - h10ageokid,
    agefbn = ifelse(agefbn < 12, NA_real_, agefbn),
  ) |> 
  mutate(.by = c(ethnicity, ragender),
    agefb = chop_equally(agefbn, 2, labels = c("Low", "High"))
  )
# TE I've plotted the mean values by year. They look reasonably stable. Beauchamp 2016 took a three year average. I don't think this is necessary

@@

# clean up cache so it doesn't get too big
rm(rand_orig, trk_orig, rands, rand_fam, rand_2010, rand_2016)
# TE we can delete more
rm(pgen,pgen4a,pgen4e,pgirep,trk_weights)

@@

correct_scores        <- readxl::read_excel("PGI_Heritabilities.xlsx", na = "NA", 
                                            range="B1:U53")

@@

map_df(1:nrow(correct_scores), function(i) {

@@

system("sed '1d;2d;3d;4d;5d;6d;7d;8d;9d' pgi_correct.res > reg_results")
reg_snp <- read.table("reg_results", header=TRUE)

@@

# Loop for twin results
map_df(1:nrow(correct_scores), function(i) {

@@

# TE TODO
#opts_chunk$set(fig.width = 12, fig.height = 8)
#570, 600
# or 5.98 and 6.79 inches

@@

# TE Note that under Fisher's fundamental theorem of natural selection, the selection
# for fitness should be equal to it's heritability... But in HRS the
# the AFB PGS much more strongly predicts RLRS than the NEB PGS and it's selection
# is much lower than what the heritability of NEB should be.
# The low R2 of NEB causes it to be removed, but the genetics for NEB could just be a little different
# in HRS versus other samples!
#NEB_result <-  corrected_effects-which(corrected_effects[,"trait"]=="NEBwomen"),]

@@

  corrected_effects$TwinH2_Source <- corrected_effects$Source

>>>>

# TE this is average number of kids that your kids have
rand$div_h10gkid <- rand$h10gkid / rand$raevbrn

# TE 283 people report having grandkids, but having had no children. Weird!
# This ugly code subsets the data by ethnicity then counts all observations
# where grandkids is greater than 0 but number of kids is 0
gkids_unusable_white <- sum(na.omit(ifelse(subset(rand, ethnicity == "white")$h10gkid > 0 & subset(rand, ethnicity == "white")$raevbrn == 0, 1, 0)))

gkids_unusable_black <- sum(na.omit(ifelse(subset(rand, ethnicity == "black")$h10gkid > 0 & subset(rand, ethnicity == "black")$raevbrn == 0, 1, 0)))

# TE if grandkids but no kids div_gkid is infinite. Replacing with NA
rand$div_h10gkid[is.infinite(rand$div_h10gkid)] <- NA

# TE added some variables for grandkid analysis
# TE I've also added sampling weights in calculation of mean values
# note the child weight is just weight* number of kids. so mean_h10gkid 
# and mean_div_h10gkid are actually the same
rand <- rand |>
  mutate(.by = rabyear, # calculate this pooling ethnicity
    mean_weight = mean(weight, na.rm = TRUE),
    mean_child_weight = mean(child_weight, na.rm = TRUE),
    mean_raevbrn = mean(raevbrn*weight/mean_weight, na.rm = TRUE),
    rlrs = raevbrn/mean_raevbrn,
    mean_h10gkid = mean(h10gkid*weight/mean_weight, na.rm = TRUE),
    rlrs_h10gkid = h10gkid/mean_h10gkid,
    mean_div_h10gkid = mean(div_h10gkid*child_weight/mean_child_weight, na.rm = TRUE), 
    rlrs_div_h10gkid = div_h10gkid/mean_div_h10gkid
  ) |>
  mutate(.by = ethnicity, # calculate these stats within-ethnicity
    income_resid = calc_income(rand, ethnicity[1]),
    income_resid = income_resid - min(income_resid, na.rm = TRUE) + 1,
    income_med = chop_equally(income_resid, 2, labels = c("Low", "High")),
    # as.numeric avoids haven.labelled issues:
    raedyrs = as.numeric(raedyrs),
    r10hearte = as.numeric(r10hearte),
    r10diabe = as.numeric(r10diabe),
    edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")),
    born = chop(rabyear, 1942, labels = c("pre-1942", "post-1942")),
    # wave 10 is 2010
    # table(rands$white$r10mstat)
     # 1                 1.married
     # 2   2.married,spouse absent
     # 3               3.partnered
     # 4               4.separated
     # 5                5.divorced
     # 6      6.separated/divorced
     # 7                 7.widowed
     # 8           8.never married
    married = factor(r10mstat == 1, 
                     levels = c(FALSE, TRUE),
                     labels = c("Other", "Married")),
    r10sibs = r10livsib + 1, # n. sibs including oneself!
    agefbn = r10agey_e - h10ageokid,
    agefbn = ifelse(agefbn < 12, NA_real_, agefbn),
  ) |> 
  mutate(.by = c(ethnicity, ragender),
    agefb = chop_equally(agefbn, 2, labels = c("Low", "High"))
  )

# TE I've plotted the mean values by year. They look reasonably stable. Beauchamp 2016 took a three year average. I don't think this is necessary

@@

# clean up cache so it doesn't get too big
rm(rand_orig, trk_orig, rands, rand_fam, rand_2010, rand_2016)

# TE we can delete more
rm(pgen,pgen4a,pgen4e,pgirep,trk_weights)

@@

####################################
# Formatting parameter values to use
####################################

# correct_scores holds h2 and R2 values, as well as useful information about the PGS
# such as group categories taken from the PGI repository and simplified names
correct_scores        <- readxl::read_excel("PGI_Heritabilities.xlsx", na = "NA", 
                                            range="B1:U53")

@@

####################################
# Calculating error corrected estimates
####################################

# Loop for for calculating results with SNP heritabilities
map_df(1:nrow(correct_scores), function(i) {

@@

# This removes "junk" rows in the way of our regression betas
system("sed '1d;2d;3d;4d;5d;6d;7d;8d;9d' pgi_correct.res > reg_results")
reg_snp <- read.table("reg_results", header=TRUE)

@@

# Loop for for calculating results with twin heritabilities
map_df(1:nrow(correct_scores), function(i) {

@@

@@

# TE Note that under Fisher's fundamental theorem of natural selection, the selection
# for fitness should be equal to it's heritability... But in HRS the
# the AFB PGS much more strongly predicts RLRS than the NEB PGS and it's selection
# is much lower than what the heritability of NEB should be.
# The low R2 of NEB causes it to be removed, but the genetics for NEB 
# could just be a little different in HRS versus other samples!
# NEB_result <-  corrected_effects-which(corrected_effects[,"trait"]=="NEBwomen"),]

@@

# Cleaning and saving corrected effects so people can have the exact figures
  corrected_effects$TwinH2_Source <- corrected_effects$Source
